---
title: "\\pkg{evinf}: An \\proglang{R} Package for Inference with Extreme Value Inflated Count Data"
plain-title: "evinf: An R Package for Inference with Extreme Value Inflated Count Data"
short-title: "evinf: Extreme Value Inflated Count Data"
author:
  - name: David Randahl
    affiliations:
      - name: Swedish Defence University
        department: Department of War Studies
        address: Drottning Kristinas Väg 47
        city: Stockholm
        country: Sweden
    email: david.randahl@fhs.se
  - name: Johan Vegelius
    affiliations:
      - name: Uppsala University
        department: Department of Medical Sciences
        city: Uppsala
        country: Sweden
abstract: |
  Count data in the social sciences are frequently generated by processes that occasionally, but not always, produce extreme values. A small number of such observations can have an outsized impact on the results of standard count regression models, and the common remedies of trimming or winsorizing the data mask an arbitrary inclusion decision rather than solving the underlying problem. The \pkg{evinf} package implements the extreme value and zero-inflated negative binomial (EVZINB) regression model and its non-zero-inflated counterpart (EVINB), which treat structural zeroes, ordinary counts, and extreme values as three separate latent states, each with its own set of covariates. The package provides routines for estimating these models by maximum likelihood, for obtaining bootstrapped and cluster-bootstrapped standard errors and $p$-values, for likelihood ratio tests of individual covariates across components, for prediction of expected values, state probabilities, and quantiles, and for comparing the fitted models with negative binomial and zero-inflated negative binomial alternatives on model fit and out-of-bag predictive performance. All model objects are compatible with the \pkg{broom} and \pkg{modelsummary} ecosystems. We describe the statistical model and its estimation, present the package interface, and illustrate its use with data on fatalities from organized violence.
keywords: [count data, extreme values, zero-inflation, negative binomial, Pareto distribution, mixture models, R]
keywords-formatted: [count data, extreme values, zero-inflation, negative binomial, Pareto distribution, mixture models, "\\proglang{R}"]
bibliography: evinf.bib
format:
  jss-pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{amsmath,amssymb}
        \newcommand{\class}[1]{`\code{#1}'}
        \newcommand{\fct}[1]{\code{#1()}}
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

<!--
DRAFT NOTES (remove before submission)
=======================================
* All code chunks are set to `eval: false` at the document level because the
  empirical example uses placeholder data blocks. Once real data are inserted,
  switch to `eval: true` (or per-chunk) and add `cache: true` on the model
  fitting chunks, which are slow.
* Placeholder data blocks are marked `## <<PLACEHOLDER>>`.
* The JSS Quarto format is provided by the quarto-journals/jss extension:
  `quarto add quarto-journals/jss`.
* Function/argument names below follow the current code in R/, not the (partly
  stale) Rd files. See dev/package_audit.md for the list of discrepancies that
  should be resolved before the vignette is finalized.
-->

# Introduction {#sec-introduction}

Count data are ubiquitous in the social sciences. Researchers model the number of fatalities from organized violence in a country-month [@eck2007one; @hultman2013united], the number of protest events in a city-week [@weidmann2019internet], the number of terrorist attacks in a year [@clauset2017enduring], the number of crimes in a neighborhood [@disha2019different], or the number of people displaced by a flood [@vestby2024societal]. For many of these phenomena, the standard modeling toolbox of Poisson and negative binomial (NB) regression, extended with zero-inflation to accommodate the structural zeroes that arise when a large share of units are simply not at risk of the outcome [@hilbe2011negative; @bagozzi2015forecasting], is adequate. Yet a common feature of exactly these phenomena is that they occasionally, but not always, produce extreme values. A conflict that has for years produced a few dozen fatalities per month may suddenly escalate into a campaign of mass killing, a protest movement may in a single week draw a hundred times more participants than it ever did before, and a river that floods a few hundred houses each year may one year displace a million people. Processes of this kind are often self-reinforcing beyond some threshold, and the resulting observations are not merely large but appear to be drawn from a different distribution than the bulk of the data.

The presence of a handful of such observations creates a well-known but rarely addressed problem for the applied researcher. Because extreme values are, by their nature, orders of magnitude larger than the vast majority of observations, they can dominate the likelihood of a count regression model and shift the estimated coefficients substantially. In @randahl2024inference, we show that removing or censoring the ten largest of nearly 20,000 country-month observations of one-sided violence against civilians in Africa, that is, the 0.05 percent most extreme values, flips the sign of the coefficient for autocracy from negative and statistically significant to positive and statistically significant and halves the coefficient for democracy. The decision to include or exclude a few observations therefore effectively decides the substantive conclusion of the analysis. The two conventional responses to this problem, to trim the largest observations from the sample or to winsorize them to some threshold value [@dixon1974trimming], do not solve it. They merely relocate the arbitrary inclusion decision to a choice of threshold, and they either discard precisely the cases that are often of greatest substantive interest or artificially alter the values of the dependent variable. Extreme values also cause practical estimation problems: NB models fitted to such data frequently fail to converge, and bootstrap or cross-validation procedures may fail for a substantial share of resamples.

The \pkg{evinf} package implements the *extreme value and zero-inflated negative binomial* (EVZINB) regression model proposed in @randahl2024inference, which addresses this problem by extending the familiar zero-inflated negative binomial (ZINB) model with a third latent state that generates extreme values. Instead of forcing the researcher to decide which observations are "too extreme" to be included, the EVZINB model treats the data as a mixture of three sub-processes, one generating structural zeroes, one generating ordinary counts, and one generating extreme values from a Pareto distribution, and estimates both the probability that an observation belongs to each state and the behavior of the outcome conditional on each state as functions of covariates. This has three benefits. First, the parameters of the count component are estimated while filtering out the influence of the extreme observations, yielding more stable and less biased estimates and fewer convergence problems. Second, the researcher can specify different covariates for each component and can therefore draw additional inferences about which factors make extreme values more likely and which factors govern how extreme they become once they occur. Third, the model produces, for every observation, both the model-implied (prior) probability of each state and the ex-post (posterior) probability of each state given the observed count, which opens up new types of analyses, for instance of the risk that a low-intensity conflict escalates into a high-intensity one. When zero-inflation is not present, the model reduces to an extreme value inflated negative binomial (EVINB) model, which is also implemented in the package.

Several \proglang{R} packages exist for count regression with excess zeroes, and several for extreme value modeling, but to the best of our knowledge none combine the two in a single regression framework. Zero-inflated and hurdle count models are available in \pkg{pscl} [@zeileis2008regression], \pkg{countreg} [@zeileis2016countreg], \pkg{glmmTMB} [@brooks2017glmmtmb], and \pkg{brms} [@burkner2017brms], among others, and NB regression in \pkg{MASS} [@venables2002modern]. Extreme value distributions can be fitted with packages such as \pkg{evd} [@stephenson2002evd], \pkg{extRemes} [@gilleland2016extremes], \pkg{ismev} [@heffernan2018ismev], and \pkg{poweRlaw} [@gillespie2015fitting], the latter implementing the discrete power-law estimation procedures of @clauset2009power. These packages are, however, built for analyzing the tail of a distribution in isolation, for example for estimating return levels or testing whether a distribution has a power-law tail, and not for jointly modeling the tail together with the body of a count distribution as functions of covariates. Finite mixture regression packages such as \pkg{flexmix} [@grun2008flexmix] allow for mixtures of count regressions, but do not include a discretized Pareto component with a data-driven threshold, nor the bootstrap-based inference that the discontinuity of the likelihood in this threshold necessitates.

This article serves as an introduction to the \pkg{evinf} package and as an expository on the statistical foundations of the EVZINB model. In @sec-model we present the model, its estimation, and the tools for inference and prediction. @sec-package describes the main functions of the package and how they fit together in a typical workflow. @sec-example illustrates the package using data on fatalities from organized violence, and @sec-guidance collects practical recommendations for model specification and estimation. @sec-conclusion concludes.

# The EVZINB model {#sec-model}

## Latent states and likelihood

The EVZINB model builds on the ZINB model, which assumes that count data are generated by two separate sub-processes, one producing zeroes and one producing counts (which may also be zero). In the ZINB model these sub-processes are modeled separately with possibly different covariates, and the model can be understood as a regression model with two latent states [@hilbe2011negative]. The EVZINB model extends this framework with a third latent state generating extreme values. We denote the three states $Z$ (zero), $NB$ (negative binomial) and $EV$ (extreme value), and define for each observation $i$ an unobserved random vector $\boldsymbol{W}_i$ which takes the value $(1,0,0)^T$, $(0,1,0)^T$, or $(0,0,1)^T$ when the observation is generated from the $Z$, $NB$, or $EV$ state, respectively. Conditional on the state, the observed count $Y_i$ has the following distribution:

$$
\begin{aligned}
Y_i \mid \boldsymbol{W}_i = (1,0,0)^T &= 0 \\
Y_i \mid \boldsymbol{W}_i = (0,1,0)^T &\sim \text{NegBin}(\mu_{NB,i}, \alpha_{NB}) \\
Y_i \mid \boldsymbol{W}_i = (0,0,1)^T &\sim \text{Pareto}(c_{EV}, \alpha_{EV,i}),
\end{aligned}
$$ {#eq-states}

where $\mu_{NB,i}$ and $\alpha_{NB}$ are the mean and dispersion parameters of the NB distribution, and $c_{EV}$ and $\alpha_{EV,i}$ are the lower bound (or threshold) and shape parameters of a Pareto distribution. With prior state probabilities $\pi_{Z,i}$, $\pi_{NB,i}$, and $\pi_{EV,i}$, summing to one, the probability mass function of $Y_i$ is

$$
f(y_i) = \pi_{Z,i}\, f_Z(y_i) + \pi_{NB,i}\, f_{NB}(y_i; \mu_{NB,i}, \alpha_{NB}) + \pi_{EV,i}\, f_{EV}(y_i; \alpha_{EV,i}, c_{EV}),
$$ {#eq-pmf}

where $f_Z(y_i)$ is one if $y_i = 0$ and zero otherwise, and $f_{NB}(\cdot)$ is the usual NB probability mass function with variance $\mu_{NB,i}(1 + \alpha_{NB}\mu_{NB,i})$, which approaches the Poisson distribution as $\alpha_{NB} \to 0$. The extreme value state is modeled with a discretized version of the Pareto distribution, obtained by integrating the continuous Pareto density over unit intervals,

$$
f_{EV}(y_i; \alpha_{EV,i}, c_{EV}) = \mathbf{1}_{y_i \geq c_{EV}} \left[ \left(\frac{c_{EV}}{y_i}\right)^{\alpha_{EV,i}} - \left(\frac{c_{EV}}{y_i + 1}\right)^{\alpha_{EV,i}} \right],
$$ {#eq-pareto}

where $\mathbf{1}_{y_i \geq c_{EV}}$ is an indicator function which is one when $y_i \geq c_{EV}$ and zero otherwise. The Pareto distribution is proportional to $y^{-\alpha_{EV}}$ and thus places substantially more probability in its tail than distributions with exponentially decaying tails, such as the NB distribution. It is also scale invariant: the conditional probability of observing a value twice as large as some $c_1 > c_{EV}$ is the same for any $c_1$ [@mandelbrot1983fractal]. This makes it a natural choice for self-reinforcing processes of the kind discussed in the introduction. The shape parameter governs how heavy the tail is; as $\alpha_{EV}$ approaches one from above, the expected value of the distribution approaches infinity, and for $\alpha_{EV} \leq 1$ the mean does not exist.

The prior state probabilities are modeled with a multinomial logistic regression with the $NB$ state as the reference category,

$$
\pi_{Z,i} = \frac{\exp\{\boldsymbol{\gamma}_Z^T \boldsymbol{x}_{\pi,i}\}}{1 + \exp\{\boldsymbol{\gamma}_Z^T \boldsymbol{x}_{\pi,i}\} + \exp\{\boldsymbol{\gamma}_{EV}^T \boldsymbol{x}_{\pi,i}\}}, \qquad
\pi_{EV,i} = \frac{\exp\{\boldsymbol{\gamma}_{EV}^T \boldsymbol{x}_{\pi,i}\}}{1 + \exp\{\boldsymbol{\gamma}_Z^T \boldsymbol{x}_{\pi,i}\} + \exp\{\boldsymbol{\gamma}_{EV}^T \boldsymbol{x}_{\pi,i}\}},
$$ {#eq-multinom}

with $\pi_{NB,i} = 1 - \pi_{Z,i} - \pi_{EV,i}$, and the conditional NB mean and the observation-specific Pareto shape are modeled with log-linear specifications,

$$
\mu_{NB,i} = \exp\{\boldsymbol{\beta}_{NB}^T \boldsymbol{x}_{NB,i}\}, \qquad
\alpha_{EV,i} = \exp\{\boldsymbol{\beta}_{EV}^T \boldsymbol{x}_{EV,i}\}.
$$ {#eq-loglinear}

Here $\boldsymbol{x}_{\pi,i}$, $\boldsymbol{x}_{NB,i}$, and $\boldsymbol{x}_{EV,i}$ are covariate vectors, each starting with a one for the intercept. For notational convenience we use the same covariates for the two multinomial equations, but the package allows different covariate sets for all four equations. The full parameter vector is thus $\boldsymbol{\theta}^T = (\boldsymbol{\gamma}_Z^T, \boldsymbol{\gamma}_{EV}^T, \boldsymbol{\beta}_{NB}^T, \boldsymbol{\beta}_{EV}^T, \alpha_{NB}, c_{EV})$, and the log likelihood is $\mathcal{L}(\boldsymbol{\theta}) = \sum_{i=1}^n \log f(y_i; \boldsymbol{\theta})$. If the $EV$ state is removed, the model reduces to the ZINB model, and if the $Z$ state is removed, it reduces to the EVINB model.

The EVZINB model can be viewed as a latent state regression model in which the states correspond to the sub-processes generating the data. In the context of organized violence, the $Z$ state corresponds to units that are not at risk of violence at all, the $NB$ state to the ordinary, moderately overdispersed count process, and the $EV$ state to episodes of mass violence. In this framing, the parameters $\boldsymbol{\gamma}_Z$ and $\boldsymbol{\gamma}_{EV}$ answer questions about *when* a unit is at risk of entering a state, and $\boldsymbol{\beta}_{NB}$ and $\boldsymbol{\beta}_{EV}$ answer questions about how the process behaves *given* its state.

## Estimation {#sec-estimation}

The parameters are estimated by maximum likelihood using a version of the expectation-maximization (EM) algorithm of @dempster1977maximum. In the E-step, the ex-post (posterior) probabilities that observation $i$ belongs to each state, commonly referred to as the responsibilities,

$$
\Pr(W_{i,j} = 1 \mid y_i) = \frac{\pi_{j,i} f_j(y_i; \boldsymbol{\theta})}{\pi_{Z,i} f_Z(y_i;\boldsymbol{\theta}) + \pi_{NB,i} f_{NB}(y_i; \boldsymbol{\theta}) + \pi_{EV,i} f_{EV}(y_i; \boldsymbol{\theta})}, \quad j \in \{Z, NB, EV\},
$$ {#eq-posterior}

are computed given the current parameter values, and the expected complete-data log likelihood $Q(\boldsymbol{\theta}; \boldsymbol{\theta}^{(0)})$ is formed. Because no closed-form solution exists for the M-step, the package uses a generalized EM (GEM) algorithm [@wu1983convergence; @lange1995gradient] in which the parameters $\tilde{\boldsymbol{\theta}} = (\boldsymbol{\gamma}_Z, \boldsymbol{\gamma}_{EV}, \boldsymbol{\beta}_{NB}, \boldsymbol{\beta}_{EV}, \alpha_{NB})$ are updated with a small number of Newton-Raphson steps, followed by a line search to guarantee that the observed log likelihood does not decrease. The threshold parameter $c_{EV}$ requires special treatment. The likelihood is discontinuous and not differentiable in $c_{EV}$, so gradient-based updates are inapplicable. However, because $\mathcal{L}(\boldsymbol{\theta})$ is monotone in $c_{EV}$ between unique observed values of $y_i$, its maximum is attained at one of the unique observed values. After each update of $\tilde{\boldsymbol{\theta}}$, $c_{EV}$ is therefore updated by evaluating the log likelihood at each candidate value in a user-specified range of the unique values of $y$ and selecting the value which maximizes it. This can be seen as a version of the expectation conditional maximization either (ECME) algorithm of @liu1994ecme. To save time when $c_{EV}$ is far from its optimum, the package first runs a warm-up phase with a small number of EM steps between each update of $c_{EV}$, and switches to full convergence runs once $c_{EV}$ has stabilized. The core routines are implemented in \proglang{C++} through \pkg{Rcpp} and \pkg{RcppArmadillo} [@eddelbuettel2011rcpp; @eddelbuettel2014rcpparmadillo]. Full details, including pseudocode, are given in the online appendix of @randahl2024inference.

## Inference {#sec-inference}

Because of the discontinuity of the likelihood in $c_{EV}$, standard errors based on the observed information matrix are not appropriate, and the package instead relies on bootstrapping for inference. The model is refitted to a large number of bootstrap resamples of the data, using the parameter estimates from the full sample as starting values, and standard errors are computed as the standard deviation of the parameter estimates across bootstraps. Bootstrapped $p$-values are obtained by inverting the bootstrap distribution of each coefficient, that is, as twice the proportion of bootstrap estimates on the opposite side of zero from the point estimate.[^pvalue] For panel or otherwise clustered data, the package supports a cluster (block) bootstrap in which all observations of a cluster are either included or excluded together, which is robust to the autocorrelation and heteroskedasticity typically present in such data [@cameron2008bootstrap]. In our experience, the bootstrap also serves a diagnostic purpose: the EVZINB model tends to be estimable on resamples where the NB model fails, and the number of failed bootstraps is itself informative about the stability of a model specification.

[^pvalue]: The package can also compute approximate $p$-values by dividing the coefficient by its bootstrapped standard error and referring the result to a $t$-distribution, which requires fewer bootstraps but relies on distributional assumptions that may not hold in the sparse extreme value component or when the residual distribution is highly skewed.

To investigate whether a covariate contributes significantly to the model as a whole, rather than to a single component, the package implements a likelihood ratio (LR) test. Let $\hat{\boldsymbol{\theta}}$ denote the unrestricted maximum likelihood estimate and $\tilde{\boldsymbol{\theta}}$ the estimate under the restriction that the coefficients of the $p$th covariate are zero in every component in which it appears. The test statistic

$$
\chi^2 = -2\left(\mathcal{L}(\tilde{\boldsymbol{\theta}}) - \mathcal{L}(\hat{\boldsymbol{\theta}})\right)
$$ {#eq-lrtest}

asymptotically follows a $\chi^2$ distribution with degrees of freedom equal to the number of restrictions, that is, the number of components in which the covariate is included. The LR test can also be carried out on each bootstrap sample, which yields a bootstrap distribution of the test statistic that can be compared with its theoretical $\chi^2$ distribution.

## Prediction {#sec-prediction}

For prediction it is common to use the expected value of the dependent variable given the covariates. In the three-state model this would be $\hat{Y} = \pi_{NB}\mu_{NB} + \pi_{EV}\mu_{EV}$, but as noted above, the mean of the Pareto distribution does not exist when $\alpha_{EV} \leq 1$, which is precisely the situation in which extreme values are most extreme. We therefore propose to replace the mean of the $EV$ state by its harmonic mean, $\tilde{\mu}_{EV} = 1/\mathbb{E}[1/Y] = c_{EV}(\alpha_{EV} + 1)/\alpha_{EV}$, which exists for any $\alpha_{EV} > 0$. The resulting point prediction is

$$
\hat{Y}_i = \pi_{NB,i}\, \exp\{\boldsymbol{\beta}_{NB}^T \boldsymbol{x}_{NB,i}\} + \pi_{EV,i}\, \frac{c_{EV}\left(\alpha_{EV,i} + 1\right)}{\alpha_{EV,i}}.
$$ {#eq-harmonic}

The harmonic mean is a conservative estimate of the expected value, and it should be kept in mind that it will understate the mean in cases where the mean exists. The package also offers an alternative prediction which replaces the Pareto mean with $\exp\{\mathbb{E}[\log Y]\} = c_{EV}\exp\{1/\alpha_{EV}\}$, the geometric mean of the Pareto distribution, which is somewhat less conservative but also exists for all $\alpha_{EV} > 0$.

Because the EVZINB distribution is heavily skewed, a single point prediction is often less informative than the quantiles of the predicted distribution. The package computes quantiles of the full three-component mixture distribution for each observation, which allows, for instance, the analyst to report the value that the model implies 5 percent of observations with a given covariate profile would exceed. In addition, the model produces the prior state probabilities $\pi_{Z,i}$, $\pi_{NB,i}$, and $\pi_{EV,i}$ for new observations, which can be used to assess the risk of a unit entering the extreme value state, and, for the estimation sample, the posterior state probabilities in @eq-posterior, which can be used to assess whether a unit, given its observed count, has in fact entered that state.

# The evinf package and main functions {#sec-package}

The \pkg{evinf} package is available from the Comprehensive \proglang{R} Archive Network (CRAN) and can be installed and loaded in the usual way.

```{r}
#| label: install
#| eval: false
install.packages("evinf")
library("evinf")
```

The package is organized around two model fitting functions, \fct{evzinb} and \fct{evinb}, which return objects of class \class{evzinb} and \class{evinb}, respectively. All other functions in the package take such an object as their first argument. The functions can be grouped into four families: functions for fitting the models, S3 methods for summarizing and presenting them, functions for prediction, and functions for testing and model comparison. @tbl-functions gives an overview.

| Function | Purpose |
|:--|:--|
| \fct{evzinb}, \fct{evinb} | Fit an EVZINB or EVINB model, with optional (cluster) bootstrap |
| \fct{print}, \fct{summary} | Print a model overview or a full summary with bootstrapped standard errors and $p$-values |
| \fct{tidy}, \fct{glance} | \pkg{broom}-style tidy coefficient tables and model fit statistics |
| \fct{coefficient\_extractor} | Extract the full set of bootstrapped coefficients for custom analyses |
| \fct{predict} | Point predictions, state probabilities, Pareto shapes, and quantiles, with optional bootstrapped intervals |
| \fct{revzinb\_fit}, \fct{revinb\_fit} | Random draws from the fitted model |
| \fct{lr\_test} | Likelihood ratio tests of covariates across components |
| \fct{compare\_models} | Fit NB and ZINB models, optionally trimmed or winsorized, on the same data and bootstraps |
| \fct{oob\_evaluation} | Out-of-bag predictive performance across bootstraps |

: Main functions of the \pkg{evinf} package. {#tbl-functions}

## Fitting models

The function \fct{evzinb} fits an EVZINB model. Its first four arguments are formulas for the four equations of the model: \code{formula\_nb} for the NB (count) component, \code{formula\_zi} for the zero-inflation component, \code{formula\_evi} for the extreme value inflation component, and \code{formula\_pareto} for the Pareto (extreme value) component. Only \code{formula\_nb} is required; the other three default to the same specification as the count component. The response variable must be the same in all formulas. In line with the recommendations discussed in @sec-guidance, we expect that most users will want to specify a more parsimonious formula for the Pareto component than for the other components.[^formulas]

[^formulas]: The formulas are used to construct the design matrices of the four components through \fct{model.frame}. In the current version of the package, this means that transformations, factor variables, and interactions should be created as columns in the data before the model is fitted rather than specified inside the formulas; see @sec-guidance.

The function estimates the model on the full sample and, by default, on \code{n\_bootstraps} bootstrap resamples of the data. The bootstrap is controlled by the arguments \code{bootstrap} (logical), \code{n\_bootstraps} (default 100, though at least 1,000 is recommended for bootstrapped $p$-values), \code{block} (the name of a cluster identifier variable in the data, which turns on the cluster bootstrap), and \code{boot\_seed} (a seed for reproducibility). Since each bootstrap requires a full refit of the model, bootstrapping is the computationally demanding part of the procedure, and the function supports parallel estimation across bootstraps through the \pkg{foreach} and \pkg{doParallel} packages by setting \code{multicore = TRUE} and, optionally, \code{ncores}. Before starting the bootstrap, the function prints a rough estimate of the expected runtime based on the time taken by the full-sample fit.

The remaining arguments control the estimation algorithm. The most important of these is \code{c.lim}, a vector of length two giving the lower and upper bounds of the range of unique observed values of the dependent variable which are considered as candidates for the threshold $c_{EV}$. The default of \code{c(50, 1000)} is appropriate for many count outcomes in the social sciences, but should be adapted to the scale of the data at hand; a warning is issued if the candidate range contains more than 100 unique values, in which case the range can be narrowed or the candidate set can be thinned using the \code{prune.c.range} argument. The convergence tolerance and the maximum number of EM steps are controlled by \code{max.diff.par} and \code{max.no.em.steps}, the number of Newton-Raphson steps per M-step by \code{no.m.bfgs.steps.*}, and the maximum parameter change per step by \code{max.upd.par.*}, where the suffixes refer to the individual components. Starting values for all parameters can be supplied through the \code{init.*} arguments, and \code{verbose = TRUE} prints the progress of the full-sample estimation and of each bootstrap.

The function \fct{evinb} fits the EVINB model without the zero-inflation component. It has the same interface as \fct{evzinb}, except that it lacks the \code{formula\_zi} argument and the arguments relating to the zero-inflation component.

## Summarizing and presenting models

The \fct{print} method gives a brief overview of a fitted model, including the four formulas, whether the full-sample estimation converged, the number of bootstraps and the number of bootstraps that failed, and the number of parameters. The \fct{summary} method returns a list of coefficient tables for the four components, together with the estimates of $\alpha_{NB}$ and $c_{EV}$ and their bootstrapped standard errors, the average state proportions, and the number of failed bootstraps. Through the arguments \code{coef} and \code{p\_value}, the user can choose whether to report the coefficients from the full-sample fit or the mean or median across bootstraps, and whether to compute bootstrapped or approximate $p$-values.

For presentation, the package provides \fct{tidy} and \fct{glance} methods following the conventions of the \pkg{broom} package [@robinson2023broom]. The \fct{tidy} method returns a tibble with one row per term, with the usual columns \code{term}, \code{estimate}, \code{std.error}, \code{statistic}, and \code{p.value}, and optionally confidence intervals, for the component selected through the argument \code{component} (one of \code{"zi"}, \code{"evi"}, \code{"count"}, \code{"pareto"}, or \code{"all"}). When \code{component = "all"}, the tibble contains an additional column \code{y.level} identifying the component, which allows \pkg{modelsummary} [@arel2022modelsummary] to produce regression tables with one column per component. The \fct{glance} method returns the number of observations and parameters, $\alpha_{NB}$, $c_{EV}$, the log likelihood, and the AIC and BIC. The package also ships a goodness-of-fit map, \code{gm\_evzinb}, which can be passed to the \code{gof\_map} argument of \fct{modelsummary} to label these statistics appropriately. For users who wish to go beyond what the methods offer, \fct{coefficient\_extractor} returns the full set of bootstrapped coefficients, one row per bootstrap, for one or all components.

## Prediction

The \fct{predict} method produces a range of predicted quantities from a fitted model, selected through the argument \code{type}. The default, \code{"harmonic"}, is the point prediction in @eq-harmonic; \code{"explog"} gives the alternative prediction based on the geometric mean of the Pareto component; \code{"counts"} gives the conditional NB mean $\mu_{NB,i}$; \code{"pareto\_alpha"} gives the observation-specific Pareto shape $\alpha_{EV,i}$; \code{"zi"}, \code{"count\_state"}, and \code{"evinf"} give the prior probabilities of the three states, which are returned jointly by \code{"states"}; and \code{"quantile"} gives the quantile of the full mixture distribution specified by the argument \code{quantile}. Finally, \code{"all"} returns all of the above in a single tibble. Predictions can be made for the estimation sample or for new data supplied through \code{newdata}. By default, predictions are based on the full-sample parameter estimates, but the argument \code{pred} can be set to \code{"bootstrap\_median"} or \code{"bootstrap\_mean"} to base them on the bootstrap distribution instead, and setting \code{confint = TRUE} returns bootstrapped confidence intervals, at the level given by \code{conf\_level}, around any of the vector-valued predictions. Setting \code{return\_bootstraps = TRUE} additionally returns the predictions from each bootstrap, which is useful for custom analyses such as the ones illustrated in @sec-example.

The functions \fct{revzinb\_fit} and \fct{revinb\_fit} draw random values from the fitted distribution for each observation in the estimation sample or in new data. These can be used for posterior predictive checks of the model against the observed distribution of the dependent variable, or for simulation-based analyses.

## Testing and comparison

The function \fct{lr\_test} carries out the LR test in @eq-lrtest for one or more covariates. Its argument \code{vars} is either a character vector of variable names, in which case each variable is tested separately if \code{single = TRUE} and all variables are restricted jointly if \code{single = FALSE}, or a list of character vectors, in which case each element of the list is tested as a joint restriction. For each test, the function refits the model with the variable(s) removed from every component in which they appear, and returns a tibble with the restricted and unrestricted log likelihoods, the degrees of freedom, the test statistic, and the $p$-value. With \code{bootstrap = TRUE}, the test is also carried out on every bootstrap sample, and the function additionally returns a Kolmogorov-Smirnov test of the bootstrapped test statistics against the theoretical $\chi^2$ distribution as well as the proportion of bootstraps in which the test is significant.

The function \fct{compare\_models} fits NB and ZINB models, using \pkg{MASS} and \pkg{pscl}, to the same data and with the same count and zero-inflation formulas as a fitted \class{evzinb} object, and refits them on exactly the same bootstrap resamples. Optionally, it also fits trimmed (\code{razorize = TRUE}) and winsorized (\code{winsorize = TRUE}) versions of these models, where the observations above the \code{cutoff\_value}th largest value are removed or censored. The result is a list of class \class{evzinbcomp} containing the original model and objects of class \class{nbboot} and \class{zinbboot}, for which the package provides \fct{tidy}, \fct{glance}, \fct{predict}, and \fct{coefficient\_extractor} methods, so that the EVZINB model and its competitors can be compared in a single table. Because the comparison models are fitted to the same bootstrap resamples, the bootstrapped model fit statistics and out-of-bag predictions of the models are paired, which allows for paired comparisons of the type reported in @randahl2024inference.

Finally, \fct{oob\_evaluation} evaluates the out-of-bag (OOB) predictive performance of a bootstrapped model, that is, the accuracy of the predictions from each bootstrap fit for the observations not included in that resample. The function returns a vector with one value per bootstrap of the root mean squared log error (the default), the root mean squared error, the mean squared error, or the mean absolute error, or of any user-supplied function of the predicted and true values.

# Using evinf: Civilian fatalities in armed conflict {#sec-example}

To illustrate the package, we model the number of civilian fatalities from organized violence at the country-year level. This outcome exhibits all three features that motivate the EVZINB model: a large proportion of structural zeroes, since most countries in most years experience no organized violence at all; a moderately overdispersed count process in countries that experience some violence; and a small number of extreme observations corresponding to genocides, mass atrocities, and wars with very high civilian tolls. The data used here follow the supplementary analysis in @randahl2024inference and are drawn from the Uppsala Conflict Data Program Georeferenced Event Dataset [UCDP GED, @sundberg2013introducing; @davies2024organized], with covariates from the Varieties of Democracy project [@coppedge2024vdem] and the World Bank.[^hks]

[^hks]: The package also bundles a reduced version of the replication data from @hultman2013united on UN peacekeeping and one-sided violence against civilians in the conflict-month, \code{data("hks", package = "evinf")}, which can be used to reproduce the replication study in @randahl2024inference.

```{r}
#| label: load-packages
library("evinf")
library("dplyr")
library("tidyr")
library("ggplot2")
library("modelsummary")
```

```{r}
#| label: load-data
## <<PLACEHOLDER: load conflict data>>
## The examples below assume a tibble `conflict_data` with one row per
## country-year and the following columns:
##   country            country identifier (used for the cluster bootstrap)
##   year               calendar year
##   civ_deaths         number of civilian fatalities (dependent variable)
##   lag_ln_civ_deaths  log(1 + civilian fatalities), previous year
##   lag_ln_comb_deaths log(1 + combatant fatalities), previous year
##   lag_ln_conflicts   log(1 + number of ongoing armed conflicts), previous year
##   lag_ln_pop         log(population), previous year
##   lag_ln_gdp         log(GDP per capita), previous year
##   lag_polyarchy      electoral democracy index (v2x_polyarchy), previous year
conflict_data <- NULL
```

## The distribution of the dependent variable

Before fitting any model, it is useful to look at the distribution of the dependent variable, both to confirm that the three-state structure is plausible and to inform the choice of the candidate range for the threshold $c_{EV}$. The code below summarizes the distribution of civilian fatalities with and without zeroes, and plots a histogram of the non-zero values on the logarithmic scale.

```{r}
#| label: dv-summary
conflict_data %>%
  summarize(
    n = n(),
    prop_zero = mean(civ_deaths == 0),
    median = median(civ_deaths),
    p90 = quantile(civ_deaths, 0.9),
    p99 = quantile(civ_deaths, 0.99),
    max = max(civ_deaths),
    overdispersion = var(civ_deaths) / mean(civ_deaths)
  )
```

```{r}
#| label: fig-dv-hist
#| fig-cap: "Distribution of non-zero civilian fatalities in the country-year data (logarithmic scale)."
conflict_data %>%
  filter(civ_deaths > 0) %>%
  ggplot(aes(x = civ_deaths)) +
  geom_histogram(bins = 40, color = "black", fill = "grey70") +
  scale_x_log10() +
  labs(x = "Civilian fatalities (log scale)", y = "Number of observations") +
  theme_minimal()
```

<!-- <<PLACEHOLDER: describe the distribution once data are in. In the ISQ
supplementary analysis: 10,246 observations, X% zeroes, unconditional
overdispersion ratio of X, a right-skewed body with a small number of
observations in the tens of thousands.>> -->

## Fitting the EVZINB model

We fit an EVZINB model in which the count, zero-inflation, and extreme value inflation components include all six covariates, while the Pareto component includes only the lagged number of civilian fatalities, the lagged number of conflicts, and lagged population. The rationale for the more parsimonious Pareto specification is that only observations above $c_{EV}$ enter into the estimation of this component, and that these observations are few; we return to this in @sec-guidance. All covariates are lagged by one year and, where appropriate, log-transformed to keep them on a narrow range. Since the data are panel data, we use a cluster bootstrap on the country level, and we run the bootstraps in parallel.

```{r}
#| label: fit-evzinb
#| cache: true
evzinb_model <- evzinb(
  formula_nb = civ_deaths ~ lag_ln_civ_deaths + lag_ln_comb_deaths +
    lag_ln_conflicts + lag_ln_pop + lag_ln_gdp + lag_polyarchy,
  formula_zi = civ_deaths ~ lag_ln_civ_deaths + lag_ln_comb_deaths +
    lag_ln_conflicts + lag_ln_pop + lag_ln_gdp + lag_polyarchy,
  formula_evi = civ_deaths ~ lag_ln_civ_deaths + lag_ln_comb_deaths +
    lag_ln_conflicts + lag_ln_pop + lag_ln_gdp + lag_polyarchy,
  formula_pareto = civ_deaths ~ lag_ln_civ_deaths + lag_ln_conflicts +
    lag_ln_pop,
  data = conflict_data,
  block = "country",
  n_bootstraps = 1000,
  boot_seed = 20260908,
  c.lim = c(50, 2000),
  multicore = TRUE,
  ncores = 8
)
evzinb_model
```

```
## <<PLACEHOLDER: print output>>
```

The printed overview shows the four formulas, whether the full-sample estimation converged, and how many of the bootstraps failed. A non-trivial number of failed bootstraps, or a full-sample fit that has not converged within the maximum number of EM steps, is a sign that the model may be mis-specified relative to the data; see @sec-guidance.

## Regression tables

The \fct{summary} method gives the complete set of results, but for presentation it is more convenient to use \fct{tidy} and \fct{modelsummary}. The \fct{tidy} method with \code{component = "all"} returns a long tibble with one row per term and component, using bootstrapped standard errors and bootstrapped $p$-values by default.

```{r}
#| label: tidy-evzinb
tidy(evzinb_model, component = "all")
```

```
## <<PLACEHOLDER: tidy output>>
```

To compare the EVZINB model with the conventional alternatives, we use \fct{compare\_models} to fit NB and ZINB models to the same data and bootstrap resamples. The ZINB model uses the count and zero-inflation formulas of the EVZINB model.[^nbfail]

[^nbfail]: In the supplementary analysis of @randahl2024inference, the NB model could not be estimated on these data because of the extreme degree of zero-inflation and overdispersion. If this happens, \code{nb\_comparison = FALSE} restricts the comparison to the ZINB model.

```{r}
#| label: compare-models
#| cache: true
model_comparison <- compare_models(
  evzinb_model,
  nb_comparison = TRUE,
  zinb_comparison = TRUE,
  multicore = TRUE,
  ncores = 8
)
```

The resulting object contains the original EVZINB model together with objects of class \class{nbboot} and \class{zinbboot}, all of which have \fct{tidy} and \fct{glance} methods. We can therefore pass them directly to \fct{modelsummary}, using the \code{shape} argument to place the components of the ZINB and EVZINB models in separate columns and the bundled \code{gm\_evzinb} goodness-of-fit map to label the model statistics. The coefficients reported here are the medians across bootstraps, and the standard errors are the standard deviations across bootstraps.

```{r}
#| label: tbl-regression
#| tbl-cap: "Regression results for the NB, ZINB, and EVZINB models. Dependent variable: civilian fatalities in the country-year. Coefficients are medians across bootstraps with bootstrapped standard errors in parentheses."
modelsummary(
  list(
    "NB" = model_comparison$nb,
    "ZINB" = model_comparison$zinb,
    "EVZINB" = model_comparison$evzinb
  ),
  component = "all",
  coef_type = "bootstrap_median",
  shape = term ~ model + y.level,
  gof_map = gm_evzinb,
  stars = c("+" = 0.1, "*" = 0.05, "**" = 0.01, "***" = 0.001)
)
```

<!-- <<PLACEHOLDER: substantive interpretation of the table. Following the
supplementary analysis in the ISQ paper: the overall pattern of effects is
similar across models, but the EVZINB model shows, for instance, that the
lagged number of civilian fatalities increases both the NB count and the
probability of entering the extreme value state, with a negative Pareto
coefficient indicating that extreme values become more extreme; and that the
lagged number of conflicts, which the ZINB model attributes solely to the
zero-inflation component, also affects the count and the extreme value
inflation.>> -->

## Likelihood ratio tests

Because the effect of a covariate is spread across four components, the individual coefficients and their $p$-values do not tell us whether a covariate contributes to the model as a whole. The LR test in @eq-lrtest addresses this question. Below we test each of the six covariates separately, and additionally test the joint contribution of the two economic covariates.

```{r}
#| label: lr-tests
#| cache: true
lr_single <- lr_test(
  evzinb_model,
  vars = c("lag_ln_civ_deaths", "lag_ln_comb_deaths", "lag_ln_conflicts",
           "lag_ln_pop", "lag_ln_gdp", "lag_polyarchy"),
  single = TRUE
)
lr_single

lr_joint <- lr_test(
  evzinb_model,
  vars = c("lag_ln_pop", "lag_ln_gdp"),
  single = FALSE
)
lr_joint
```

```
## <<PLACEHOLDER: LR test output>>
```

The degrees of freedom of each test equal the number of components in which the covariate appears, so that the three covariates included in the Pareto component are tested with four degrees of freedom and the remaining three with three.

## State probabilities and predicted fatalities

The main advantage of the EVZINB model over the ZINB model is that it separates the effect of a covariate on the risk of a unit entering the extreme value state from its effect on the ordinary count process and on the extremeness of the extreme values. The \fct{predict} method makes these different quantities available. Below we construct a grid of hypothetical observations in which the lagged number of conflicts varies over its observed range while all other covariates are held at their means, and predict the prior state probabilities, the harmonic mean prediction, and the 50th and 95th percentiles of the predicted distribution for each point on the grid. For the point prediction we also request bootstrapped 90 percent confidence intervals.

```{r}
#| label: predict-grid
#| cache: true
prediction_grid <- conflict_data %>%
  summarize(across(starts_with("lag_"), mean)) %>%
  select(-lag_ln_conflicts) %>%
  crossing(lag_ln_conflicts = seq(0, log1p(20), length.out = 50))

state_probs <- predict(evzinb_model, newdata = prediction_grid, type = "states")

harmonic_pred <- predict(
  evzinb_model,
  newdata = prediction_grid,
  type = "harmonic",
  confint = TRUE,
  conf_level = 0.9
)

q50_pred <- predict(evzinb_model, newdata = prediction_grid,
                    type = "quantile", quantile = 0.5)
q95_pred <- predict(evzinb_model, newdata = prediction_grid,
                    type = "quantile", quantile = 0.95)

predictions <- prediction_grid %>%
  bind_cols(state_probs, harmonic_pred) %>%
  mutate(q50 = q50_pred, q95 = q95_pred,
         n_conflicts = expm1(lag_ln_conflicts))
```

```{r}
#| label: fig-state-probs
#| fig-cap: "Predicted prior probabilities of the zero, count, and extreme value states as a function of the lagged number of ongoing armed conflicts, with all other covariates held at their means."
predictions %>%
  select(n_conflicts, pr_zc, pr_count, pr_pareto) %>%
  pivot_longer(-n_conflicts, names_to = "state", values_to = "probability") %>%
  mutate(state = factor(state, levels = c("pr_zc", "pr_count", "pr_pareto"),
                        labels = c("Zero", "Count", "Extreme value"))) %>%
  ggplot(aes(x = n_conflicts, y = probability)) +
  geom_line() +
  facet_wrap(~ state) +
  labs(x = "Number of ongoing conflicts (lag)", y = "State probability") +
  theme_minimal()
```

```{r}
#| label: fig-predictions
#| fig-cap: "Predicted civilian fatalities (harmonic mean, with 90 percent bootstrapped confidence interval) and the predicted 50th and 95th percentiles of civilian fatalities as a function of the lagged number of ongoing armed conflicts, with all other covariates held at their means."
predictions %>%
  select(n_conflicts, harmonic, ci_lb, ci_ub, q50, q95) %>%
  pivot_longer(c(harmonic, q50, q95), names_to = "quantity", values_to = "value") %>%
  mutate(quantity = factor(quantity, levels = c("harmonic", "q50", "q95"),
                           labels = c("Predicted fatalities",
                                      "50th percentile", "95th percentile"))) %>%
  ggplot(aes(x = n_conflicts, y = value)) +
  geom_ribbon(data = . %>% filter(quantity == "Predicted fatalities"),
              aes(ymin = ci_lb, ymax = ci_ub), alpha = 0.2) +
  geom_line() +
  facet_wrap(~ quantity, scales = "free_y") +
  scale_y_continuous(trans = "log1p") +
  labs(x = "Number of ongoing conflicts (lag)", y = "Civilian fatalities") +
  theme_minimal()
```

<!-- <<PLACEHOLDER: interpretation. In the ISQ supplementary analysis, the
probability of the zero state drops rapidly as the number of lagged conflicts
increases from zero, while the probability of the extreme value state
increases with the number of conflicts. Contrast the median and the 95th
percentile: covariates may have divergent effects on the most common case and
on the more extreme cases.>> -->

The distinction between the 50th and the 95th percentile is worth emphasizing. In @randahl2024inference we show that the presence of UN military troops reduces the median number of civilian killings in a conflict-month but does not reduce, and for moderate troop levels increases, the 95th percentile, because it both lowers the probability of the extreme value state and makes the extreme values more extreme once they occur. Divergent effects of this kind are invisible in the NB and ZINB models, whose single count process forces every covariate to shift the entire predicted distribution in the same direction.

For observations in the estimation sample, the posterior state probabilities in @eq-posterior are stored in the fitted object and can be used to identify which observations the model assigns to the extreme value state given their observed counts. The code below lists the country-years with the highest posterior probability of belonging to the extreme value state.

```{r}
#| label: posterior-states
conflict_data %>%
  mutate(posterior_pareto = evzinb_model$fitted$posterior_pareto,
         prior_pareto = evzinb_model$fitted$prob_pareto) %>%
  arrange(desc(posterior_pareto)) %>%
  select(country, year, civ_deaths, prior_pareto, posterior_pareto) %>%
  head(10)
```

```
## <<PLACEHOLDER: posterior state output>>
```

## Model fit and predictive performance

Finally, we compare the EVZINB model with the NB and ZINB models on model fit and OOB predictive performance. The \fct{glance} method gives the log likelihood, AIC, and BIC of the full-sample fits.

```{r}
#| label: glance-models
bind_rows(
  NB = glance(model_comparison$nb),
  ZINB = glance(model_comparison$zinb),
  EVZINB = glance(model_comparison$evzinb),
  .id = "model"
)
```

```
## <<PLACEHOLDER: glance output>>
```

Because the comparison models are fitted to the same bootstrap resamples as the EVZINB model, we can also compare the models pairwise across bootstraps. Below we compute the OOB root mean squared log error of the EVZINB model with \fct{oob\_evaluation}, extract the corresponding statistic which \fct{compare\_models} stores for each bootstrap of the NB and ZINB models, and report the median OOB error of each model as well as the proportion of bootstraps in which the EVZINB model outperforms each competitor.

```{r}
#| label: oob-comparison
#| cache: true
oob_evzinb <- oob_evaluation(evzinb_model, metric = "rmsle")
oob_zinb <- purrr::map_dbl(model_comparison$zinb$bootstraps,
                           ~ if (inherits(.x, "try-error")) NA else .x$oob_rmsle)
oob_nb <- purrr::map_dbl(model_comparison$nb$bootstraps,
                         ~ if (inherits(.x, "try-error")) NA else .x$oob_rmsle)

tibble(evzinb = oob_evzinb, zinb = oob_zinb, nb = oob_nb) %>%
  summarize(
    across(everything(), ~ median(.x, na.rm = TRUE), .names = "median_{.col}"),
    prop_evzinb_better_zinb = mean(evzinb < zinb, na.rm = TRUE),
    prop_evzinb_better_nb = mean(evzinb < nb, na.rm = TRUE),
    n_failed_nb = sum(is.na(nb))
  )
```

```
## <<PLACEHOLDER: OOB output>>
```

As a last check of the model, we can compare the distribution of the observed dependent variable with the distribution of random draws from the fitted model, obtained with \fct{revzinb\_fit}. If the model is a good description of the data-generating process, the quantiles of the simulated data should track the quantiles of the observed data across the full range of the distribution, including the tail.

```{r}
#| label: posterior-predictive
set.seed(20260908)
simulated <- revzinb_fit(evzinb_model, n_draws = 100)

probs <- c(0.5, 0.75, 0.9, 0.95, 0.99, 0.999)
tibble(
  p = probs,
  observed = quantile(conflict_data$civ_deaths, probs),
  simulated = purrr::map_dbl(probs, ~ median(purrr::map_dbl(simulated, quantile, .x)))
)
```

```
## <<PLACEHOLDER: posterior predictive output>>
```

# Practical guidance {#sec-guidance}

The EVZINB model adds two regression equations to the ZINB model, and it estimates the Pareto equation on the small subset of observations above $c_{EV}$. It is therefore more demanding of the data than the models it extends, and some care in specification is required. We collect our recommendations here.

**Be parsimonious in the Pareto component.** Only observations above $c_{EV}$ contribute to the estimation of $\boldsymbol{\beta}_{EV}$, and in most applications these are a small fraction of the sample. We recommend including in the Pareto component only covariates for which there is a clear theoretical expectation that they affect how extreme the extreme values become, and excluding covariates which do not vary among the observations above $c_{EV}$. Dummy variables which are almost always one above the threshold, such as an indicator of any violence in the previous period, are a typical example of the latter. Because the covariates are exponentiated in @eq-loglinear and the range of plausible Pareto shapes is narrow [@clauset2009power], we also recommend log-transforming or standardizing continuous covariates in this component, or using covariates which operate on a narrow range.

**Choose the candidate range for the threshold deliberately.** The argument \code{c.lim} determines which unique values of the dependent variable are considered as candidates for $c_{EV}$. If the lower bound is set too low, the extreme value state will absorb observations from the body of the count distribution; if it is set too high, the Pareto component will be estimated on too few observations. A histogram of the non-zero values of the dependent variable, as in @fig-dv-hist, is usually sufficient to identify a plausible range, and the estimated $c_{EV}$ should not lie at either bound of the range. If the range contains many unique values, estimation slows down since the log likelihood is evaluated at every candidate in every EM iteration; in this case the range can be narrowed or the candidate set thinned with \code{prune.c.range}.

**Treat non-convergence as information.** Non-convergence and unstable parameter estimates should, in our view, be interpreted as a sign that the EVZINB model is either mis-specified or unsuitable for the data. This may be because a covariate in the Pareto component has little or no effect on the shape parameter, because the data-generating process is not a three-state mixture, or because there is too little data in one of the components. When no extreme value state is present in the data, the intercept of the extreme value inflation equation is pushed towards minus infinity, which is easily detected in the output and indicates that a ZINB model should be used instead. Similarly, when no zero-inflation is present, the EVINB model is the appropriate choice.

**Prepare covariates before fitting.** In the current version of the package, the design matrices are constructed directly from the variables named in the formulas. Transformations such as logarithms, squared terms, factor variables, and interactions should therefore be created as columns in the data before fitting, and referred to by name in the formulas, as in the example above.

**Plan for the bootstrap.** Bootstrapped $p$-values require a large number of bootstraps, and we recommend at least 1,000. Since each bootstrap is a full refit, this can take a considerable amount of time on large data. Fitting the model with \code{bootstrap = FALSE} first is a cheap way to check convergence and the estimated $c_{EV}$ before committing to the bootstrap, and the runtime estimate printed by \fct{evzinb} gives a rough indication of how long the bootstrap will take. For clustered data, the \code{block} argument should always be used.

# Conclusion {#sec-conclusion}

This paper has introduced the \pkg{evinf} package for \proglang{R}, which implements the extreme value and zero-inflated negative binomial regression model and its non-zero-inflated counterpart. The model addresses a problem that is common but rarely dealt with in a principled manner in applied count regression: that a small number of extreme observations may dominate the results, and that the conventional remedies of trimming and winsorizing merely relocate an arbitrary decision. By treating structural zeroes, ordinary counts, and extreme values as three separate latent states with their own covariates, the EVZINB model allows the researcher to keep all observations in the analysis while drawing separate inferences about what makes extreme values likely and what makes them extreme.

The package provides a complete workflow for such analyses, from estimation with bootstrapped inference, through presentation with \pkg{broom} and \pkg{modelsummary}, to prediction of expected values, state probabilities, and quantiles, likelihood ratio tests across components, and comparison with negative binomial and zero-inflated negative binomial models on model fit and out-of-bag predictive performance. Through an example on civilian fatalities in armed conflict, we have illustrated how these tools can be combined to draw inferences that are not available from conventional count models, in particular about the divergent effects of covariates on the most common and the most extreme cases.

We see several avenues for future development. A natural extension is to allow the threshold $c_{EV}$ to depend on covariates, which would make it possible to ask what factors affect the point at which a process may transition from a well-behaved count process to an extreme value process. Another is to embed the latent states in a hidden Markov structure for panel data, so that the transition between states can itself be modeled over time. We hope that the package will encourage researchers to revisit research questions on data with extreme values which have previously been set aside because of the modeling problems that such data cause.

# Computational details {.unnumbered}

The results in this paper were obtained using \proglang{R} `r paste(R.version$major, R.version$minor, sep = ".")` with the \pkg{evinf} `r packageVersion("evinf")` package. \proglang{R} itself and all packages used are available from CRAN at <https://CRAN.R-project.org/>.

# Acknowledgments {.unnumbered}

<!-- <<PLACEHOLDER: funding (ERC ViEWS 694640, RJ M21-0002, VR 2017-01175 in
the ISQ paper) and acknowledgments.>> -->

# References {.unnumbered}

::: {#refs}
:::
